---
title: "GeoMorpho"
format: html
editor: visual
---

# Why Study Wing Veins? - Pen version

```{r include=FALSE}
#working directory
#knitr::opts_knit$set(root.dir = "C:\Users\meloe\OneDrive\School\Biology\Bio 455\My code")
```

```{r}
#libraries
library(StereoMorph)
library(tidyverse)
library(rgl)
library(geomorph)
library(RRPP)
library(ggplot2)
library(Morpho)
library(randomForest)
library(caret)
library(abind)
library(plotly)
library(dawai)

```

If you're scared, go look at the README. Hopefully that'll clarify things!

Un-comment out the code the code block below to run stereomorph

```{r}
#| label: Landmarking client Stereomorph
#digitizeImage(image.file='wings/images', shapes.file="wings/Shapes",landmarks.ref="wings/landmarks.txt")

```

Sex separated code for Eumel- for morphoJ (outputs tps)

```         
#Sex separated code for all specimens and separating to just Eumel
#read in all metadata
wingdata = read_csv("wings/wings_final_eumel.csv") #

#split by sex
wings_m = wingdata %>% filter(Sex == "M")
wings_f = wingdata %>% filter(Sex == "F")

#converting the groups to factors
wings_m$Label<-as.factor(wings_m$Label)
wings_f$Label<-as.factor(wings_f$Label)


#create list of male and female files
male_ids = wings_m$Accession
male_files <- paste0("wings/shapes/JRW", male_ids, ".txt") #need to change cause this is a lazy workaround

female_ids = wings_f$Accession
female_files <- paste0("wings/shapes/JRW", female_ids, ".txt")

#get shapes for each group
male_shapes <- readShapes(male_files)
male_data <- readland.shapes(male_shapes)

female_shapes <- readShapes(female_files) #turns stereomorph files into a list
female_data <- readland.shapes(female_shapes)



land_mat = as.matrix(male_data$landmarks)
land_mat_again = male_shapes$landmarks.pixel

fdir <- "C:/Users/meloe/OneDrive/School/Biology/Bio 455/My code/wings/Shapes_eumel" #change where this goes

writeLMToTPS(fdir, 'Shapes_eumel.tps')
```

Sex separated code for all specimens with Eumel toggle

```{r}
#Sex separated code for all specimens
#read in all metadata
wingdata = read_csv("wings/wings_final_all.csv") 

#wingdata= wingdata %>% filter(Subgenus == 'Eumel') 
#if you put Eumel sorting back, make sure to comment out the anova and cva for subgenus

#split by sex
wings_m = wingdata %>% filter(Sex == "M")
wings_f = wingdata %>% filter(Sex == "F")

#converting the groups to factors
wings_m$Label<-as.factor(wings_m$Label)
wings_f$Label<-as.factor(wings_f$Label)
wings_m$Subgenus<-as.factor(wings_m$Subgenus)
wings_f$Subgenus<-as.factor(wings_f$Subgenus)

#create list of male and female files
male_ids = wings_m$Accession
male_files <- paste0("wings/shapes/JRW", male_ids, ".txt") #need to change cause JRW is a lazy workaround

female_ids = wings_f$Accession
female_files <- paste0("wings/shapes/JRW", female_ids, ".txt")

#get shapes for each group
male_shapes <- readShapes(male_files)
male_data <- readland.shapes(male_shapes)

female_shapes <- readShapes(female_files) #turns stereomorph files into a list
female_data <- readland.shapes(female_shapes)

#procrustes fits
MY.gpa <- gpagen(male_data, print.progress = F)
MY.gpa = rotate.coords(MY.gpa, "flipY")


FY.gpa <- gpagen(female_data, print.progress = F)
FY.gpa = rotate.coords(FY.gpa, "flipY")
#male pca
MPCA <- gm.prcomp(MY.gpa$coords)

#summary(MPCA)
plot(MPCA,label=TRUE)

mpc_scores<-MPCA[[10]]

#female pca
FPCA <- gm.prcomp(FY.gpa$coords)

#summary(FPCA)
plot(FPCA,label=TRUE)

fpc_scores<-FPCA[[10]]


#preparing for visualization
mpc_scores2<-cbind(wings_m,mpc_scores)
fpc_scores2<-cbind(wings_f,fpc_scores)


#scree plots for weight of principal components
mscree = MPCA$sdev^2 / sum(MPCA$sdev^2)
qplot(c(1:34), mscree) + #34 is the number of principal components of the pca, check by looking at length of scree
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Male Bee Principal Component Importance") +
  ylim(0, 1)

fscree = FPCA$sdev^2 / sum(FPCA$sdev^2)
qplot(c(1:34), fscree) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Female Bee Principal Component Importance") +
  ylim(0, 1)


#ggplots
ggplot(mpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Male Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) # + geom_text(label=mpc_scores2$Label, vjust=2)

ggplot(fpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Female Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) #+ geom_text(label=fpc_scores2$Label, vjust=2)



anov_label_m = procD.lm(MY.gpa$coords~wings_m$Label)
#plot(anov_label_m)
#summary(anov_label_m)

anov_label_f = procD.lm(FY.gpa$coords~wings_f$Label)
#summary(anov_label_f)

anov_subgen_m = procD.lm(MY.gpa$coords~wings_m$Subgenus)
#summary(anov_subgen_m)

anov_subgen_f = procD.lm(FY.gpa$coords~wings_f$Subgenus)
#summary(anov_subgen_f)


cvaa_label_m = CVA(MY.gpa$coords, groups = mpc_scores2$Label, cv=TRUE)
#cvaa_subgenus_m = CVA(MY.gpa$coords, groups = mpc_scores2$Subgenus, cv=TRUE)

cvaa_label_m
#cvaa_subgenus_m

cvaa_label_f = CVA(FY.gpa$coords,groups = fpc_scores2$Label, cv=TRUE)
cvaa_subgenus_f = CVA(FY.gpa$coords,groups = fpc_scores2$Subgenus, cv=TRUE)

cvaa_label_f
cvaa_subgenus_f
```

```{r}
#Sex separated code for all specimens and separating to just Eumel
#read in all metadata
wingdata = read_csv("wings/wings_final_eumel.csv") #wing_asap has columns Accession, Sex, Label, ASAP, and Sequence

#wingdata_eumel = wingdata %>% filter(Subgenus == 'Eumel') 

#split by sex
wings_m = wingdata %>% filter(Sex == "M")
wings_f = wingdata %>% filter(Sex == "F")

#converting the groups to factors
wings_m$Label<-as.factor(wings_m$Label)
wings_f$Label<-as.factor(wings_f$Label)


#create list of male and female files
male_ids = wings_m$Accession
male_files <- paste0("wings/shapes/JRW", male_ids, ".txt") #need to change cause this is a lazy workaround

female_ids = wings_f$Accession
female_files <- paste0("wings/shapes/JRW", female_ids, ".txt")

#get shapes for each group
male_shapes <- readShapes(male_files)
male_data <- readland.shapes(male_shapes)

female_shapes <- readShapes(female_files) #turns stereomorph files into a list
female_data <- readland.shapes(female_shapes)

#procrustes fits
MY.gpa <- gpagen(male_data, print.progress = F)

MY.gpa = rotate.coords(MY.gpa, "flipY")


FY.gpa <- gpagen(female_data, print.progress = F)
FY.gpa = rotate.coords(FY.gpa, "flipY")
#male pca
MPCA <- gm.prcomp(MY.gpa$coords)

#summary(MPCA)
plot(MPCA,label=TRUE)

mpc_scores<-MPCA[[10]]

#female pca
FPCA <- gm.prcomp(FY.gpa$coords)

#summary(FPCA)
plot(FPCA,label=TRUE)

fpc_scores<-FPCA[[10]]

#preparing for visualization
mpc_scores2<-cbind(wings_m,mpc_scores)
fpc_scores2<-cbind(wings_f,fpc_scores)

#scree plots for weight of principal components
mscree = MPCA$sdev^2 / sum(MPCA$sdev^2)
qplot(c(1:34), mscree) + #34 is the number of principal components of the pca, check by looking at length of scree
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Male Bee Principal Component Importance") +
  ylim(0, 1)

fscree = FPCA$sdev^2 / sum(FPCA$sdev^2)
qplot(c(1:34), fscree) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Female Bee Principal Component Importance") +
  ylim(0, 1)

#ggplots
ggplot(mpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Eumelissodes Male Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75)  #+ geom_text(label=mpc_scores2$Accession, vjust=2)

ggplot(mpc_scores2, aes(x=Comp2,y=Comp3, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Eumelissodes Male Geometric Morphometrics",
       x = "PC2", y = "PC3") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) # + geom_text(label=mpc_scores2$Label, vjust=2)

ggplot(mpc_scores2, aes(x=Comp1,y=Comp3, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Eumelissodes Male Geometric Morphometrics",
       x = "PC1", y = "PC3") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) # + geom_text(label=mpc_scores2$Label, vjust=2)



ggplot(fpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Eumelissodes Female Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) #+ geom_text(label=fpc_scores2$Accession, vjust=2)

ggplot(fpc_scores2, aes(x=Comp3,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Eumelissodes Female Geometric Morphometrics",
       x = "PC3", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) #+ geom_text(label=fpc_scores2$Label, vjust=2)

anov_label_m = procD.lm(MY.gpa$coords~wings_m$Label)
#plot(anov_label_m)
summary(anov_label_m)

anov_label_f = procD.lm(FY.gpa$coords~wings_f$Label)
summary(anov_label_f)

cvaa_label_m=CVA(MY.gpa$coords,groups = mpc_scores2$Label, cv=TRUE)
cvaa_label_f=CVA(FY.gpa$coords,groups = fpc_scores2$Label, cv=TRUE)

cvaa_label_m
cvaa_label_f
```

All specimens (with Eumel option) unseparated by sex

```{r}
#unseparated by sex
#read in all metadata
wingdata = read_csv("wings/wings_final_all.csv") #wing_asap has columns Accession, Sex, Label, ASAP, and Sequence

#wingdata = wingdata %>% filter(Subgenus == 'Eumel') 


wingdata$Label<-as.factor(wingdata$Label)
wingdata$Sex<-as.factor(wingdata$Sex)
wingdata$Subgenus <- as.factor(wingdata$Subgenus)



wing_ids = wingdata$Accession
wing_files = paste0("wings/shapes/JRW", wing_ids, ".txt")
wing_shapes <- readShapes(wing_files)
wing_data <- readland.shapes(wing_shapes)



Y.gpa <- gpagen(wing_data, print.progress = F)
Y.gpa = rotate.coords(Y.gpa, "flipY")
#plotAllSpecimens(Y.gpa$coords, label = TRUE)
#plotOutliers(Y.gpa$coords, inspect.outliers = TRUE)
#the only male Agilis4 is our sole outlier



anov_sex = procD.lm(Y.gpa$coords~wingdata$Sex)
summary(anov_sex)


anov_label = procD.lm(Y.gpa$coords ~ wingdata$Label)
summary(anov_label)

#anov_genus = procD.lm(Y.gpa$coords ~ wingdata$Subgenus )
#summary(anov_genus)

PCA <- gm.prcomp(Y.gpa$coords)

#summary(PCA)
#plot(PCA,label=TRUE)
pc_scores<-PCA[[10]]
pc_scores2<-cbind(wingdata,pc_scores)


cvaa_label=CVA(Y.gpa$coords,groups = pc_scores2$Label, cv=TRUE)
cvaa_sex=CVA(Y.gpa$coords,groups = pc_scores2$Sex, cv=TRUE)
#cvaa_subgenus = CVA(Y.gpa$coords,groups = pc_scores2$Subgenus, cv=TRUE)

cvaa_label
cvaa_sex
#cvaa_subgenus


ggplot
ggplot(pc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label, shape=Sex), size=3) + 
  labs(title = "All Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) #+ 
  #geom_text(label=pc_scores2$Label, vjust=2)

ggplot(pc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "All Geometric Morphometrics",
     x = "PC1", y = "PC2") +
 theme_classic() + stat_ellipse(type="norm", level = 0.75) + 
 #geom_text(label=pc_scores2$Label, vjust=2) +
 facet_wrap(Sex~., nrow=2)

```

Everything below is outdated

```{r include = FALSE}
#read in all metadata
wingmetadata = read_csv("wings/wings_asap_pen_copy_copy.csv") #wing_asap has columns Accession, Sex, Label, ASAP, and Sequence
asap_labels = read_csv("wings/ASAP_label_pen.csv") #asap_label has columns ASAP and ASAP label
wingdata = wingmetadata %>% left_join(asap_labels, by = "ASAP") #squishing them together i think

#split by sex
wings_m = wingdata %>% filter(Sex == "M")
wings_f = wingdata %>% filter(Sex == "F")

#converting the groups to factors
wings_m$Label<-as.factor(wings_m$Label)
wings_f$Label<-as.factor(wings_f$Label)


#create list of male and female files
male_ids = wings_m$Accession
male_files <- paste0("wings/shapes/JRW", male_ids, ".txt") #need to change cause this is a lazy workaround

female_ids = wings_f$Accession
female_files <- paste0("wings/shapes/JRW", female_ids, ".txt")

#get shapes for each group
male_shapes <- readShapes(male_files)
male_data <- readland.shapes(male_shapes)

female_shapes <- readShapes(female_files) #turns stereomorph files into a list
female_data <- readland.shapes(female_shapes)


```

## Normalizing Landmarks (Image Alignment)

*Ryan and Kaiona's code autogenerated missing landmarks. Don't do that. There was only one sample with a missing landmark, so I took it out of the dataset. - Pen*

```{r}
#procrustes fits
MY.gpa <- gpagen(male_data, print.progress = F)

MY.gpa = rotate.coords(MY.gpa, "flipY")


FY.gpa <- gpagen(female_data, print.progress = F)
FY.gpa = rotate.coords(FY.gpa, "flipY")
```

### Variation of normalized male landmarks

*This allows us to visualize outliers, and confirm that landmarks have been placed correctly. Not essential for biological analysis, but helpful nonetheless - Pen*

```{r echo=FALSE}
plotAllSpecimens(MY.gpa$coords, label = TRUE)

plotOutliers(MY.gpa$coords, inspect.outliers = TRUE)
```

### Variation of normalized female landmarks

```{r echo=FALSE}
plotAllSpecimens(FY.gpa$coords, label = TRUE)

plotOutliers(FY.gpa$coords, inspect.outliers = TRUE)
```

By visualizing the landmarks this way, we can confirm that there were no issues with data entry, as specimens with incorrectly-placed landmarks would be detected as outliers at this point.

## Initial PCA of landmarks

```{r}
#male pca
MPCA <- gm.prcomp(MY.gpa$coords)

#summary(MPCA)
plot(MPCA,label=TRUE)

mpc_scores<-MPCA[[10]]

#female pca
FPCA <- gm.prcomp(FY.gpa$coords)
#summary(FPCA)
plot(FPCA,label=TRUE)

fpc_scores<-FPCA[[10]]

#preparing for visualization
mpc_scores2<-cbind(wings_m,mpc_scores)
fpc_scores2<-cbind(wings_f,fpc_scores)

#scree plots for weight of principal components
mscree = MPCA$sdev^2 / sum(MPCA$sdev^2)
qplot(c(1:34), mscree) + #34 is the number of principal components of the pca, check by looking at length of scree
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Male Bee Principal Component Importance") +
  ylim(0, 1)

fscree = FPCA$sdev^2 / sum(FPCA$sdev^2)
qplot(c(1:34), fscree) + 
  geom_line() + 
  xlab("Principal Component") + 
  ylab("Variance Explained") +
  ggtitle("Female Bee Principal Component Importance") +
  ylim(0, 1)

#ggplots
ggplot(mpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Male Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75)  + geom_text(label=mpc_scores2$Label, vjust=2)

ggplot(fpc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  geom_point(aes(color=Label), size=3) + 
  labs(title = "Female Geometric Morphometrics",
       x = "PC1", y = "PC2") +
  theme_classic() + stat_ellipse(type="norm", level = 0.75) + geom_text(label=fpc_scores2$Label, vjust=2)
```

These visualizations plot each specimens' values for the first two principal components, colors according to the group we assigned them going in, then draws an ellipse around all the points in a group with at least 3 specimens. We can see that the males are all extremely clustered. Individuals in the agilis/trinodis complex can be approximately differentiated from individuals of other species based on the value of PC1, but only barely, and individual variation is still huge. Female bees clustered similarly, for the most part, but individuals of the same group can still be placed very far from one another.

```{r}
anov_label = procD.lm(MY.gpa$coords~wings_m$Label)


plot(anov_label)
summary(anov_label)
```

### Combining all shape data, departitioning by sex.

```{r}
wingdata$Label<-as.factor(wingdata$Label)
wingdata$Sex<-as.factor(wingdata$Sex)

wing_ids = wingdata$Accession
wing_files = paste0("wings/shapes/JRW", wing_ids, ".txt")
wing_shapes <- readShapes(wing_files)
wing_data <- readland.shapes(wing_shapes)



Y.gpa <- gpagen(wing_data, print.progress = F)
Y.gpa = rotate.coords(Y.gpa, "flipY")
#plotAllSpecimens(Y.gpa$coords, label = TRUE)
#plotOutliers(Y.gpa$coords, inspect.outliers = TRUE)
#the only male Agilis4 is our sole outlier

PCA <- gm.prcomp(Y.gpa$coords)
#summary(PCA)
#plot(PCA,label=TRUE)
pc_scores<-PCA[[10]]

#pc_scores2<-cbind(wingdata,pc_scores)

#ggplot
#ggplot(pc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
  #geom_point(aes(color=Label, shape=Sex), size=3) + 
  #labs(title = "All Geometric Morphometrics",
  #     x = "PC1", y = "PC2") +
 # theme_classic() + stat_ellipse(type="norm", level = 0.75) 
 #+ geom_text(label=pc_scores2$Label, vjust=2)

#ggplot(pc_scores2, aes(x=Comp1,y=Comp2, color=Label)) + 
#  geom_point(aes(color=Label), size=3) + 
#  labs(title = "All Geometric Morphometrics",
  #     x = "PC1", y = "PC2") +
 # theme_classic() + stat_ellipse(type="norm", level = 0.75) + 
 #+ geom_text(label=pc_scores2$Label, vjust=2) +
 # facet_wrap(Sex~., nrow=2)
```

```{r}
anov_sex = procD.lm(Y.gpa$coords~wingdata$Sex)
anov_label = procD.lm(MY.gpa$coords~wings_m$Label)

plot(anov_sex)
```
